A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-05-01 paper.tex
Fixes to section 2 & 3

\newtheorem*{exmp}{Example}

@@

\begin{exmp}
We introduce a running example to assist the explanation. We consider a
simple arbiter system in which the environment requests for either 1 or 2
resources, and the controller may grant access to two resources. The total
number of requests grows each round by the number of environment requests and
shrinks by the number of resources granted by the controller in the previous
round.  The controller must ensure that the number of unhandled requests does
not accumulate to more than 2.  Figure~\ref{fig:example} shows the variables
(\ref{fig:examplevars}) and the formulas for computing next-state variables
(\ref{fig:exampletrans}) for this example. We use primed identifiers to denote
next-state variables and curly braces to define the domain of a variable.

This example is the $n=2$ instance of the more general problem of an arbiter of
$n$ resources. The set of winning states grows exponentially with the number of
resources the controller could use to grant requests, which makes the problem
hard for BDD solvers. In Section~3 we will outline how this example can be
solved without enumerating all winning states.

@@

The game abstraction is initially constructed from an opponent candidate
strategy. It is refined by counterexamples to the solver's own candidate
strategy that are formed when the opponent solves the abstract game
corresponding to that strategy. By abstracting the game in this way we
conceptually direct the solvers' search to consider opponent actions that may
lead to winning strategies.

@@

    \begin{subfigure}[t]{.1\textwidth}

@@

                            node [left=4pt] {\texttt{gr0=0 gr1=1}}

@@

@@

@@

    \begin{subfigure}[t]{.3\textwidth}

@@

                            edge from parent [fixed] node [below left=0pt and -5pt] {\texttt{req=2}}
                            edge from parent [fixed] node [above left=-3pt and 0pt] {\texttt{req=2}}

@@

                node [above=4pt] {$\langle s, 2 \rangle$};

@@

    To solve our example, the algorithm begins by invoking a SAT solver to find
    a trace of length $k$ (here we consider $k=3$) from the initial state to
    the error state (see Figure~\ref{fig:trace}). This trace is used to create
    an abstract game tree by taking the first action in the trace and fixing it
    as the first action in the game tree (Figure~\ref{fig:agt}).  This abstract
    game tree represents the conjecture that \texttt{req=1} may be a good first
    action.  Next the solver attempts to find a counterexample to this
    conjecture by finding a trace that avoids the error state with the first
    environment action fixed to \texttt{req=1} (Figure~\ref{fig:trace2}), i.e.
    by playing the abstract game on behalf of the controller.

    Observe that the controller chooses to grant only one resource, just enough
    to ensure that an error state will not be reached. This example will show
    the worst case when the SAT solver returns actions that will lead to large
    abstract game trees. However, in practice we employ heuristics to guide the
    SAT solver away from the worst case.

    The environment's candidate is refined using the controller's
    counterexample traces. In Figure~\ref{fig:agt2}, the game tree tree is
    refined based on the first action of the controller's trace. The
    \texttt{req=1} action in the second round is still winning under the
    assumption that the controller will not grant any resources. The
    environment is free to make that assumption because the controller's
    counterexample does not yet contain any grants. Eventually, the refinements
    to the environment's candidate will force the controller to grant more
    resources and result in a game tree such as in Figure~\ref{fig:refined}.

The bounded synthesis algorithm solves abstract game trees by invoking a SAT
solver to search for candidate partial strategies. Candidate strategies are
used as game abstractions for the opponent and can be verified by a call to the
opponent solver. The solver recursively refines the abstraction by finding
counterexample strategies. The full procedure is illustrated in
Algorithm~\ref{alg:bounded}.

@@

The algorithm takes a concrete game $G$ with maximum bound $\kappa$ as an implicit
argument.  In addition, it takes a player $p$ (controller or environment),
state $s$, bound $k$ and an abstract game tree $T$ and returns a winning
partial strategy for $p$, if one exists.  The initial invocation of the
algorithm takes the initial state $I$, bound $\kappa$ and an empty abstract
game tree $\emptyset$.  Initially the solver is playing on behalf of the
environment since that player takes the first move in every game round.  The
empty game tree does not constrain opponent moves, hence solving such an
abstraction is equivalent to solving the original concrete game.

@@

%%%\begin{exmp}
%%%    The controller's partial strategy in
%%%    Figure~\ref{fig:strategy} is defeated by the counterexample action
%%%    \texttt{req=2}.  Figure~\ref{fig:refined} shows a refined AGT that includes
%%%    this action.
%%%\end{exmp}

@@

We instead present an approach that iteratively solves games of increasing
bound while learning winning states from abstract games using interpolation. We
show that reaching a fixed point in learned states is sufficient for
completeness.

@@

We extend the bounded synthesis algorithm to learn states losing for one of the
players from failed attempts to find candidate strategies.  The learning
procedure kicks in whenever \textsc{findCandidate} cannot find a candidate
strategy for an abstract game tree $T$ with initial state $s$ and bound $k$.
For the controller, this means the environment partial strategy represented by $T$
will always reach $E$ from $s$ no matter the assignments to $C$ variables (the
bound is irrelevant).  For the environment, the controller can avoid $E$ for $k$
rounds no matter the assignments to $U$ variables. Hence, we have established
that $s$ is a losing state either always (for the controller) or for bound $k$ (for
the environment. We can learn additional losing states from the tree via
interpolation.  This is achieved in lines~18--20 in
Algorithm~\ref{alg:bounded}, enabled in the unbounded version of the algorithm,
which invoke \textsc{learn} or \textsc{\textoverline{learn}} to learn
controller or environment losing states respectively
(Algorithm~\ref{alg:learn}).

@@

        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left,text width=1cm] {\texttt{gr0=1, gr1=0}}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                    }
                    node [left=5pt] {$n$}
                    edge from parent [fixed] node [left,text width=1cm] {\texttt{gr0=0, gr1=1}}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                    }
                    edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \caption{A losing AGT $T$}
        \label{fig:interpolatetree}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    node [left=5pt] {$n$}
                    edge from parent [fixed] node [left,text width=1cm] {\texttt{gr0=1, gr1=0}}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                    }
                    edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_1$}
        \label{fig:treef1}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] {}
                child {node [circle,draw,right=5pt] {}
                    edge from parent [fixed] node [left,text width=1cm] {\texttt{gr0=1, gr1=0}}
                }
                child {node [circle,draw,left=5pt] {}
                    edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                }
                node [left=5pt] {$n$};
        \end{tikzpicture}
        \caption{Tree slice $T_2$}
        \label{fig:treef2}
    \end{subfigure}
    \caption{Splitting of an abstract game tree by the learning procedure.}
    \label{fig:interpolanttrees}
\end{figure}

@@

    Consider the game tree $T$ in Figure~\ref{fig:interpolatetree} from our
    running example.  This tree is labelled with controller actions that
    prevent the environment from forcing the game into the error state. No
    matter how many requests the environment makes, the controller will grant a
    surplus. So the \textsc{findCandidate} function can not find a candidate
    strategy for the environment.  We now learn the states for which these
    controller actions can force the game to stay in the controller winning
    set.

@@

    In the pictured trees the interpolant $\texttt{nrequests} = 0$ is reachable
    at $n$ and no move the environment chooses will reach an error state in one
    round from this set. A BDD solver using an uncontrollable predecessor
    operation on the error set would be forced to find a much more complicated
    set: $\texttt{nrequests} = 3 \lor (\texttt{nrequests} = 2 \land (
    \texttt{resource0} = 0 \lor \texttt{resource1} = 0)) \lor
    (\texttt{nrequests} = 1 \land (\texttt{resource0} = 0 \land
    \texttt{resource1} = 0))$. This set gets more complicate with more
    resources in the system. The advantage of interpolation is that we may
    overapproximate the losing states.

>>>>

\newtheorem*{exmp}{Example}
\newtheorem*{exmpI}{Example: Intuition behind the algorithm}
\newtheorem*{exmpInt}{Example: Why we use interpolants}

@@

We introduce a running example to assist the explanation. We consider a simple
arbiter system in which the environment makes a request for a number of
resources (1 or 2), and the controller may grant access to up to two resources.
The total number of requests grows each round by the number of environment
requests and shrinks by the number of resources granted by the controller in
the previous round.  The controller must ensure that the number of unhandled
requests does not accumulate to more than 2.  Figure~\ref{fig:example} shows
the variables (\ref{fig:examplevars}) and the formulas for computing next-state
variables (\ref{fig:exampletrans}) for this example. We use primed identifiers
to denote next-state variables and curly braces to define the domain of a
variable.

This example is the $n=2$ instance of the more general problem of an arbiter of
$n$ resources. The set of winning states has no compact representation, which
makes the problem hard for BDD solvers. In Section~3 we will outline how this
example can be solved without enumerating all winning states.

@@

@@

    \begin{subfigure}[t]{.15\textwidth}

@@

@@

                node [left=4pt] {\texttt{gr0=1 gr1=1}}

@@

                        edge from parent [fixed] node [left] {\texttt{req=2}}

@@

@@

@@

@@

\begin{exmpI}

    %% Make it clear that this is intuition only

    To solve our example, the algorithm begins by invoking a SAT solver to find
    a trace of length $k$ (here we consider $k=3$) from the initial state to an
    error state (see Figure~\ref{fig:trace}).  Given a trace we make an initial
    conjecture that the first action, \texttt{req=1}, is winning for the
    environment. This conjecture is captured in the abstract game tree show in
    Figure~\ref{fig:agt}.  Next the solver attempts to find a counterexample to
    this conjecture by finding a trace that avoids the error states with the
    first environment action fixed to \texttt{req=1}, i.e. by playing the
    abstract game on behalf of the controller.

    The resulting trace is shown in Figure~\ref{fig:trace2}. The environment's
    candidate is refined using this trace and Figure~\ref{fig:agt2} shows the
    refined game tree.  The controller's counterexample was to grant two
    resources in the first round, to which the environment responds by now
    making two requests in the second round.

    Eventually, the refinements to the environment's candidate will force the
    controller to grant more resources and result in a game tree such as in
    Figure~\ref{fig:refined}.

\end{exmpI}

@@

The full procedure is illustrated in Algorithm~\ref{alg:bounded}.  The
algorithm takes a concrete game $G$ with maximum bound $\kappa$ as an implicit
argument.  In addition, it takes a player $p$ (controller or environment),
state $s$, bound $k$ and an abstract game tree $T$ and returns a winning
partial strategy for $p$, if one exists.  The initial invocation of the
algorithm takes the initial state $I$, bound $\kappa$ and an empty abstract
game tree $\emptyset$.  Initially the solver is playing on behalf of the
environment since that player takes the first move in every game round.  The
empty game tree does not constrain opponent moves, hence solving such an
abstraction is equivalent to solving the original concrete game.

@@

@@

We instead present an approach that iteratively solves games of increasing
bound while learning bad states from abstract games using interpolation. We
show that reaching a fixed point in learned states is sufficient for
completeness.

@@

We extend the bounded synthesis algorithm to learn states losing for one of the
players from failed attempts to find candidate strategies.  The learning
procedure kicks in whenever \textsc{findCandidate} cannot find a candidate
strategy for an abstract game tree. We can learn additional losing states from
the tree via interpolation.  This is achieved in lines~18--20 in
Algorithm~\ref{alg:bounded}, enabled in the unbounded version of the algorithm,
which invoke \textsc{learn} or \textsc{\textoverline{learn}} to learn
controller or environment losing states respectively
(Algorithm~\ref{alg:learn}).

%%%For the controller, this means the environment partial strategy represented by $T$
%%%will always reach $E$ from $s$ no matter the assignments to $C$ variables (the
%%%bound is irrelevant).  For the environment, the controller can avoid $E$ for $k$
%%%rounds no matter the assignments to $U$ variables. Hence, we have established
%%%that $s$ is a losing state either always (for the controller) or for bound $k$
%%%(for the environment).

@@

        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left=2pt,text width=1cm] {\texttt{gr0=1 gr1=0}}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right=2pt,text width=1cm] {\texttt{gr0=1 gr1=1}}
                    }
                    node [left=5pt] {$n$}
                    edge from parent [fixed] node [left=2pt,text width=1cm] {\texttt{gr0=0 gr1=1}}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right=2pt,text width=1cm] {\texttt{gr0=1 gr1=1}}
                    }
                    edge from parent [fixed] node [right=2pt,text width=1cm] {\texttt{gr0=1 gr1=1}}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \caption{A losing AGT $T$}
        \label{fig:interpolatetree}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    node [left=5pt] {$n$}
                    edge from parent [fixed] node [left=2pt,text width=1cm] {\texttt{gr0=1 gr1=0}}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right=2pt,text width=1cm] {\texttt{gr0=1 gr1=1}}
                    }
                    edge from parent [fixed] node [right=2pt=2pt=2pt=2pt,text width=1cm] {\texttt{gr0=1 gr1=1}}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_1$}
        \label{fig:treef1}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] {}
                child {node [circle,draw,right=5pt] {}
                    edge from parent [fixed] node [left=2pt,text width=1cm] {\texttt{gr0=1 gr1=0}}
                }
                child {node [circle,draw,left=5pt] {}
                    edge from parent [fixed] node [right=2pt,text width=1cm] {\texttt{gr0=1 gr1=1}}
                }
                node [left=5pt] {$n$};
        \end{tikzpicture}
        \caption{Tree slice $T_2$}
        \label{fig:treef2}
    \end{subfigure}
    \caption{Splitting of an abstract game tree by the learning procedure.}
    \label{fig:interpolanttrees}
\end{figure}

@@

\begin{exmpInt}

    Consider node $n$ in Figure~\ref{fig:interpolatetree}. At this node there
    are two controller actions that prevent the environment from forcing the
    game into an error state in one game round. We want to use this tree to
    learn the states from which the controller can win playing one of these
    actions.

    One option is using a BDD solver, working backwards from the error set, to
    find all losing states. One iteration of this operation on our example
    would give the set: $\texttt{nrequests} = 3 \lor (\texttt{nrequests} = 2
    \land ( \texttt{resource0} = 0 \lor \texttt{resource1} = 0)) \lor
    (\texttt{nrequests} = 1 \land (\texttt{resource0} = 0 \land
    \texttt{resource1} = 0))$.  In the general case there is no compact
    representation of the losing set, so we try to avoid computing it by
    employing interpolation instead. The potential benefit of interpolation is
    that an interpolant can approximate the losing states in such a way that
    there is just enough information to prove realisability. It is also cheap
    to compute an interpolant from the proof of unsatisfiability of a SAT
    solver.

    At node $n$, the interpolant $\texttt{nrequests} = 1 \land
    \texttt{resource1} = 1$ captures the information we need. Any action by the
    environment followed by one of the controller actions at $n$ will be
    winning for the controller.

\end{exmpInt}

@@
